---
title: "Initial method evaluation"
author: "Joyce Hsiao"
date: "2019-02-20"
output: 
  workflowr::wflow_html:
      toc: true
      float: true
---

## Introduction

* Experimental data: PBMC of 2,683 samples and ~ 11,000 genes, including 7+ cell types. This data has large number of zeros (93% zeros in the count matrix).

* Simulation parameters
  + p = 1000 genes
  + N = 90 samples total, 45 samples per group
  + 100 simulated datasets per scenario
  
* Simulation steps:
  1. Draw (sample without replacement) p genes and N samples from the experimental dataset. 
  2. Assign samples to group 1 or 2 across genes

* Methods included: edgeR, DESeq2, glm poisson, glm quasipoisson, limma + voom, MAST, t-test, wilcoson test


## Preprocessing 

```{r}
knitr::opts_chunk$set(warning=F, message=F)
```

extract dsc output and get p-values, q-values, true signals, etc.

```{r, warning=F, message=F, eval=F}
get_pvals <- function(dir_dsc, dsc_files, verbose = F) {
  dsc_files$method <- as.factor(dsc_files$method)
  n_methods <- nlevels(dsc_files$method)

  res <- vector("list",n_methods)
  for (i in 1:nrow(dsc_files)) {
    if (verbose) {print(i)}
    fl_pval <- readRDS(file.path(dir_dsc,
                         paste0(as.character(dsc_files$method.output.file[i]), ".rds")))
    fl_beta <- readRDS(file.path(dir_dsc,
                     paste0(as.character(dsc_files$data_poisthin.output.file[i]), ".rds")))
    seed <- dsc_files$data_poisthin.seed[i]
    prop_null <- dsc_files$data_poisthin.prop_null[i]
    fl_qval <- readRDS(file.path(dir_dsc,
                        paste0(as.character(dsc_files$qvalue.output.file[i]), ".rds")))
    res[[i]] <- data.frame(method = as.character(dsc_files$method)[i],
                           seed = seed,
                           pval = fl_pval$pval,
                           qval = fl_qval$qval,
                           truth_vec = fl_beta$beta != 0,
                           prop_null = prop_null,
                           stringsAsFactors = F)
  }
  res_merge <- do.call(rbind, res)
  return(res_merge)
}
  


library(dscrutils)
dir_dsc <- "/scratch/midway2/joycehsiao/dsc-log-fold-change/benchmark"
dsc_res <- dscquery(dir_dsc, 
                    targets=c("data_poisthin", 
                              "data_poisthin.seed", 
                              "data_poisthin.prop_null",
                              "method",
                              "qvalue"))
saveRDS(dsc_res, file = "output/eval_initial.Rmd/dsc_res.rds")


pvals_res <- get_pvals(dir_dsc, dsc_res)
saveRDS(pvals_res, file = "output/eval_initial.Rmd/pvals_res.rds")
```

Some plotting and summary functions

```{r}
# type I error related functions ----------
plot_oneiter_pval <- function(pvals_res_oneiter, cols, seed=1, bins=30) {
    n_methods <- length(unique(pvals_res_oneiter$method))
    print(
    ggplot(pvals_res_oneiter, aes(x=pval, fill=method)) +
            facet_wrap(~method) +
            geom_histogram(bins=bins) +
#            xlim(xlims[1],xlims[2]) +
            scale_fill_manual(values=cols)  )
}

plot_oneiter_qq <- function(pvals_res_oneiter, cols, plot_overlay=T,
                   title_label=NULL, xlims=c(0,1), pch.type="S") {
    methods <- unique(pvals_res_oneiter$method)
    n_methods <- length(methods)
    
    if(plot_overlay) {
    print(
    ggplot(pvals_res_oneiter, aes(sample=pval, col=method)) +
            stat_qq(cex=.7) +
            scale_color_manual(values=cols)  )
    } else {
    print(
    ggplot(pvals_res_oneiter, aes(sample=pval, col=method)) +
            facet_wrap(~method) +
            stat_qq(cex=.7) +
            scale_color_manual(values=cols)  )
    }
}

# power related functions ----------

get_roc_est <- function(pvals_res, fpr_nbin=100) {
    method_list <- levels(factor(pvals_res$method))
    seed_list <- unique(pvals_res$seed)
    
    out_roc_est <- lapply(1:length(method_list), function(i) {
      df_sub <- pvals_res %>% filter(method==method_list[i] & prop_null==prop_null)
      roc_est_seed <- lapply(1:length(seed_list), function(j) {
        roc_set_seed_one <- with(df_sub[df_sub$seed==seed_list[j],],
                                 pROC::auc(response=truth_vec, predictor=qval))
        fpr <- 1-attr(roc_set_seed_one, "roc")$specificities
        tpr <- attr(roc_set_seed_one, "roc")$sensitivities
        data.frame(fpr=fpr,tpr=tpr,seed=seed_list[j])
      })
      roc_est_seed <- do.call(rbind, roc_est_seed)
      fpr_range <- range(roc_est_seed$fpr)
      fpr_seq <- seq.int(from=fpr_range[1], to = fpr_range[2], length.out = fpr_nbin+1)
      tpr_est_mean <- rep(NA, fpr_nbin)
      for (index in 1:fpr_nbin) {
        tpr_est_mean[index] <- mean( roc_est_seed$tpr[which(roc_est_seed$fpr >= fpr_seq[index] & roc_est_seed$fpr < fpr_seq[index+1])], na.rm=T) 
      }
      fpr_bin_mean <- fpr_seq[-length(fpr_seq)]+(diff(fpr_seq)/2)
      roc_bin_est <- data.frame(fpr_bin_mean=fpr_bin_mean,tpr_est_mean=tpr_est_mean)
      roc_bin_est <- roc_bin_est[!is.na(roc_bin_est$tpr_est_mean),]
      roc_bin_est$method <- method_list[i]
      return(roc_bin_est)
    }) 
    out <- do.call(rbind, out_roc_est)
    out$method <- factor(out$method)
    return(out)
}

```


## Type I error between methods

```{r, fig.width=12}
library(tidyverse)

pvals_res <- readRDS("output/eval_initial.Rmd/pvals_res.rds")

plot_type1 <- function(pvals_res, alpha) {
  n_methods <- length(unique(pvals_res$method))
  cols <- RColorBrewer::brewer.pal(n_methods,name="Dark2")
  pvals_res %>% filter(prop_null==1) %>% 
    group_by(method, seed) %>%
    summarise(type1=mean(pval<alpha, na.rm=T), nvalid=sum(!is.na(pval))) %>%
    ggplot(., aes(x=method, y=type1, col=method)) +
        geom_hline(yintercept=alpha, 
                color = "red", size=.5) +
        geom_boxplot(width=.5) +
      ylab("Type I error") +
      scale_x_discrete(position = "top",
                       labels=c("deseq2", "edger","glm_p", "glm_q",
                                "limma_v", "mast", "t_test", "wilcox")) +
      scale_color_manual(values=cols) + ggtitle(paste0("Type I error at alpha < ",alpha))
}

library(cowplot)
plot_grid(plot_type1(pvals_res, alpha=.01), 
          plot_type1(pvals_res, alpha=.001))
```

## p-value distribution

```{r, fig.width=8}
n_methods <- length(unique(pvals_res$method))
cols <- RColorBrewer::brewer.pal(n_methods,name="Dark2")
plot_grid(plot_oneiter_pval(pvals_res[pvals_res$seed == 2 & pvals_res$prop_null == 1, ],
                  cols=cols) + ylim(0,30),
          plot_oneiter_qq(pvals_res[pvals_res$seed == 2 & pvals_res$prop_null == 1, ],
                cols=cols, plot_overlay = T) + xlim(-3,-1) + ylim(0,.25) )
```


## Power: Mean AUC 

```{r, fig.width=6, eval=T}
library(pROC)

pvals_res <- readRDS("output/eval_initial.Rmd/pvals_res.rds")

plot_roc <- function(roc_est, cols,
                     title_label=NULL) {
  n_methods <- length(unique(roc_est$method))
  print(
    ggplot(roc_est, aes(x=fpr_bin_mean, 
                        y=tpr_est_mean, col=method)) +
      # geom_hline(yintercept=alpha, 
      #         color = "red", size=.5) +
      geom_step() +
      scale_color_manual(values=cols) 
    )
}


# AUC ----------
plot_auc <- function(pvals_res) {
  n_methods <- length(unique(pvals_res$method))
  cols <- RColorBrewer::brewer.pal(n_methods,name="Dark2")
  pvals_res %>% filter(prop_null != 1) %>%
    group_by(method, prop_null,seed) %>%
    summarise(auc_est=roc(response=truth_vec, predictor=qval)$auc) %>%
    group_by(method, prop_null) %>%
    summarise(auc_mean=mean(auc_est)) %>%
    ggplot(., aes(x=method, y=auc_mean, col=method, shape=factor(prop_null))) +
      geom_point(size=4) +
      ylab("Mean AUC") +
      scale_color_manual(values=cols) +
      scale_x_discrete(position = "top",
                       labels=c("deseq2", "edger", "glm_p", "glm_q", 
                                "limma_v", "mast", "t_test", "wilcox"))
}
plot_auc(pvals_res)
```

## Power: ROC 

```{r}
# get estimated ROC
roc_est <- rbind(cbind(get_roc_est(pvals_res[pvals_res$prop_null==.9,]),prop_null=.9),
                 cbind(get_roc_est(pvals_res[pvals_res$prop_null==.5,]),prop_null=.5))

roc_est %>% group_by(prop_null) %>%
  ggplot(., aes(x=fpr_bin_mean, 
                y=tpr_est_mean, col=method)) +
    facet_wrap(~prop_null) +
    geom_line() +
    scale_color_manual(values=cols) + xlim(0,.2) + ylim(0,.2) +
    geom_vline(xintercept=.05, col="gray50") +
    xlab("False positive rate ") +
    ylab("Mean true positive rate")
```


## FDR control at .01

```{r}
fdr_thres <- .01
n_methods <- length(unique(pvals_res$method))
cols <- RColorBrewer::brewer.pal(n_methods,name="Dark2")
pvals_res %>% group_by(method, seed, prop_null) %>%
  filter(prop_null != 1) %>%
  summarise(fdr_control = sum(qval < fdr_thres & truth_vec==F, na.rm=T)/sum(truth_vec==F)) %>%
  ggplot(., aes(x=method, y=fdr_control, col=method)) +
    facet_wrap(~prop_null) +
    geom_violin() +
    scale_color_manual(values=cols) 
```
