---
title: "Initial method evaluation"
author: "Joyce Hsiao"
date: "2019-04-22"
output: 
  workflowr::wflow_html:
      toc: true
      float: true
---

## Introduction/summary

* Determine relationship between effect size, sample size and power. Choose parameters of the true effect distribution accordingly. 

* Experimental data: PBMC of 2,683 samples and ~ 11,000 genes, including 7+ cell types. This data has large number of zeros (93% zeros in the count matrix).

* Simulation parameters
  + number of genes: 1,000 randomly sampled from experimental data
  + number of samples per group: (50, 50), (100, 100), (250, 250); draw n1+n2 from experimental data, then randomly assigned to group 1 or group 2
  + fraction of true effects: .1
  + distribution of true effects: normal distribution with mean 0 and sd .5, 1, 2, 4
  


## Extract dsc results

```{r}
knitr::opts_chunk$set(warning=F, message=F)

library(dscrutils)
library(tidyverse)
```

extract dsc output and get p-values, q-values, true signals, etc.

```{r, warning=F, message=F, eval=F}
dir_dsc <- "/scratch/midway2/joycehsiao/dsc-log-fold-change/pipe_power_choose_betasd"

dsc_res_edger <- dscquery(dir_dsc, 
                    targets=c("data_poisthin_choose_betasd",
                              "data_poisthin_choose_betasd.seed", 
                              "data_poisthin_choose_betasd.n1",
                              "data_poisthin_choose_betasd.betasd",
                              "edger", "pval_rank"), 
                    ignore.missing.file = T)

dsc_res_limma_voom <- dscquery(dir_dsc, 
                    targets=c("data_poisthin_choose_betasd",
                              "data_poisthin_choose_betasd.seed", 
                              "data_poisthin_choose_betasd.n1",
                              "data_poisthin_choose_betasd.betasd",
                              "limma_voom", "pval_rank"), 
                    ignore.missing.file = T)


dsc_res_edger$output.file <- dsc_res_edger$edger.output.file
dsc_res_limma_voom$output.file <- dsc_res_limma_voom$limma_voom.output.file

dsc_res_edger <- subset(dsc_res_edger, select = -c(edger.output.file))
dsc_res_limma_voom <- subset(dsc_res_limma_voom, select = -c(limma_voom.output.file))
  
dsc_res <- rbind(data.frame(dsc_res_edger, method = rep("edger", nrow(dsc_res_edger))),
                 data.frame(dsc_res_limma_voom, method = rep("limma_voom", nrow(dsc_res_limma_voom))) )


method_vec <- as.factor(dsc_res$method)
n_methods <- nlevels(method_vec)

res <- vector("list",n_methods)
for (i in 1:nrow(dsc_res)) {
#  if (verbose) {print(i)}
  fl_pval <- readRDS(file.path(dir_dsc,
                       paste0(as.character(dsc_res$output.file[i]), ".rds")))
  fl_beta <- readRDS(file.path(dir_dsc,
                   paste0(as.character(dsc_res$data_poisthin_choose_betasd.output.file[i]), ".rds")))
  seed <- dsc_res$data_poisthin_choose_betasd.seed[i]
  n1 <- dsc_res$data_poisthin_choose_betasd.n1[i]
  betasd <- dsc_res$data_poisthin_choose_betasd.betasd[i]
  fl_qval <- readRDS(file.path(dir_dsc,
                      paste0(as.character(dsc_res$pval_rank.output.file[i]), ".rds")))
  res[[i]] <- data.frame(method = as.character(dsc_res$method)[i],
                         seed = seed,
                         betasd=betasd,
                         n1=n1,
                         truth_vec = fl_beta$beta != 0,
                         pval = fl_pval$pval,
                         qval = fl_qval$qval,
                         stringsAsFactors = F)
}
res_merge <- do.call(rbind, res)

saveRDS(res_merge, file = "output/eval_initial.Rmd/res_merge.rds")
```


## Analyze

```{r}
res_merge <- readRDS(file = "output/eval_initial.Rmd/res_merge.rds")

fdr_thres <- .1

p1 <- res_merge %>%
  filter(method == "limma_voom") %>%
  group_by(seed, betasd, n1) %>%
  summarise(power = sum(truth_vec==TRUE & qval<fdr_thres)/sum(truth_vec==TRUE)) %>%
  ggplot(., aes(x=factor(n1), y=power, col=factor(betasd))) + ylim(0,1) +
    geom_boxplot() +
    ylab("Power") + xlab("n1") +
    ggtitle("limma_voom")

p2 <- res_merge %>%
  filter(method == "edger") %>%
  group_by(seed, betasd, n1) %>%
  summarise(power = sum(truth_vec==TRUE & qval<fdr_thres)/sum(truth_vec==TRUE)) %>%
  ggplot(., aes(x=factor(n1), y=power, col=factor(betasd))) + ylim(0,1) +
    geom_boxplot() +
    ylab("Power") + xlab("n1") +
    ggtitle("edger")

library(cowplot)
plot_grid(p1,p2)
```



