---
title: "Initial method evaluation: type I error"
author: "Joyce Hsiao"
date: "2019-04-22"
output: 
  workflowr::wflow_html:
      toc: true
      float: true
---

## Introduction

* Evaluate type I error of some DE methods for data with potential confounded design

* Assume equal library size for all samples

* Experimental data: PBMC of 2,683 samples and ~ 11,000 genes, including 7+ cell types. This data has large number of zeros (93% zeros in the count matrix).

* Simulation parameters
  + number of genes: 1,000 randomly sampled from experimental data
  + number of samples per group: (50, 50), (250, 250); draw n1+n2 from experimental data, then randomly assigned to group 1 or group 2
  + fraction of true effects: .1
  + distribution of true effects: normal distribution with mean 0 and sd 1
  

---

## Extract dsc results

```{r}
knitr::opts_chunk$set(warning=F, message=F)

library(dscrutils)
library(tidyverse)
```

extract dsc output and get p-values, q-values, true signals, etc.

```{r, warning=F, message=F, eval=F}
dir_dsc <- "/scratch/midway2/joycehsiao/dsc-log-fold-change/pipe_type1"

dsc_res <- dscquery(dir_dsc, 
                    targets=c("data_poisthin_null",
                              "data_poisthin_null.seed", 
                              "data_poisthin_null.n1",
                              "method", "pval_rank"), 
                    ignore.missing.file = T)

method_vec <- as.factor(dsc_res$method)
n_methods <- nlevels(method_vec)


res <- vector("list",n_methods)
for (i in 1:nrow(dsc_res)) {
#  print(i)
  fl_pval <- readRDS(file.path(dir_dsc,
                       paste0(as.character(dsc_res$method.output.file[i]), ".rds")))
  # fl_beta <- readRDS(file.path(dir_dsc,
  #                  paste0(as.character(dsc_res$data_poisthin_null.output.file[i]), ".rds")))
  seed <- dsc_res$data_poisthin_null.seed[i]
  n1 <- dsc_res$data_poisthin_null.n1[i]
  # fl_qval <- readRDS(file.path(dir_dsc,
  #                    paste0(as.character(dsc_res$pval_rank.output.file[i]), ".rds")))
  res[[i]] <- data.frame(method = as.character(dsc_res$method)[i],
                         seed = seed,
                         n1=n1,
                         pval = fl_pval$pval,
                         stringsAsFactors = F)
}
res_merge <- do.call(rbind, res)

saveRDS(res_merge, file = "output/eval_initial_type1.Rmd/res_merge.rds")
```



## Analysis


```{r, fig.width=12}
res_merge <- readRDS(file = "output/eval_initial_type1.Rmd/res_merge.rds")

make_plots <- function(res, alpha, labels,
                       args=list(n1, labels)) {
  n_methods <- length(unique(res$method))
  cols <- RColorBrewer::brewer.pal(n_methods,name="Dark2")
  res %>% filter(n1==args$n1) %>% 
    group_by(method, seed) %>%
    summarise(type1=mean(pval<alpha, na.rm=T), nvalid=sum(!is.na(pval))) %>%
#    group_by(method) %>%
    # summarise(mn=mean(type1, na.rm=T), 
    #           n=sum(!is.na(type1)), se=sd(type1, na.rm=T)/sqrt(n)) %>%
    ggplot(., aes(x=method, y=type1, col=method)) +
        # geom_errorbar(aes(ymin=mn+se, ymax=mn-se), width=.3) + 
        geom_boxplot() + geom_point(size=.7) + xlab("") +
      ylab("Type I error") +
      scale_x_discrete(position = "top",
                       labels=args$labels) +
      scale_color_manual(values=cols) +
      theme(axis.text.x=element_text(angle = 20, vjust = -.3, hjust=-.1))
}

library(cowplot)
levels(factor(res_merge$method))
labels <- c("deseq2", "edger", "limma_v", "sva_voom", "t_log2cpm_q", "wilcoxon")

# title <- ggdraw() + draw_label("Type I error at alpha < .001 \n n: 50/group",
#                                fontface='bold')
p1 <- make_plots(res_merge, alpha=.001,
                 args=list(n1=50, labels=labels)) + 
                ggtitle("Type error at alpha < .001, 50/group") + #ylim(0,.02) +
                geom_hline(yintercept=.001, col="gray30", lty=3)
p1

p2 <- make_plots(res_merge, alpha=.001,
                 args=list(n1=250, labels=labels)) + 
                ggtitle("Type error at alpha < .001, 250/group") + #ylim(0,.02) +
                geom_hline(yintercept=.001, col="gray30", lty=3)
p2
```


