---
title: "Method eval applying to bulk RNA-seq: type I error"
author: "Joyce Hsiao"
date: "2019-04-29"
output: 
  workflowr::wflow_html:
      toc: true
      float: true
---

## Introduction

* Evaluate type I error of some DE methods, using default normalization and filtering steps: edger, deseq2, limma_voom, t_test + input log2(Y+1), t_test + input log2CPM expression data quantiled normalized per gene, wilcoxon + input count data

* Assume equal library size for all samples

* Experimental data: GTEx V6 lung tissue, 320 samples and 16,069 genes.

* Simulation parameters
  + number of genes: 1,000 randomly sampled from experimental data
  + number of samples per group: (5,5), (10,10), (50,50), (150,150); draw n1+n2 from experimental data, then randomly assigned to group 1 or group 2
  + fraction of true effects: 0
  + distribution of true effects: normal distribution with mean 0 and sd 1
  
  

---

## Extract dsc results

```{r}
knitr::opts_chunk$set(warning=F, message=F)

library(dscrutils)
library(tidyverse)
```

extract dsc output and get p-values, q-values, true signals, etc.

```{r, warning=F, message=F, eval=F}
dir_dsc <- "/scratch/midway2/joycehsiao/dsc-log-fold-change/pipe_gtex"

dsc_res <- dscquery(dir_dsc, 
                    targets=c("data_poisthin_gtex",
                              "data_poisthin_gtex.seed", 
                              "data_poisthin_gtex.n1",
                              "data_poisthin_gtex.prop_null",
                              "method", "pval_rank"), 
                    ignore.missing.file = T)

method_vec <- as.factor(dsc_res$method)
n_methods <- nlevels(method_vec)
dsc_res <- dsc_res[dsc_res$method != "sva_limma_voom" & dsc_res$method != "sva_ttest",]
res <- list()
for (i in 1:nrow(dsc_res)) {
  print(i)
  fl_pval <- readRDS(file.path(dir_dsc,
                       paste0(as.character(dsc_res$method.output.file[i]), ".rds")))
  fl_beta <- readRDS(file.path(dir_dsc,
                    paste0(as.character(dsc_res$data_poisthin_gtex.output.file[i]), ".rds")))
  prop_null <- dsc_res$data_poisthin_gtex.prop_null[i]
  seed <- dsc_res$data_poisthin_gtex.seed[i]
  n1 <- dsc_res$data_poisthin_gtex.n1[i]
  # fl_qval <- readRDS(file.path(dir_dsc,
  #                    paste0(as.character(dsc_res$pval_rank.output.file[i]), ".rds")))
  res[[i]] <- data.frame(method = as.character(dsc_res$method)[i],
                         seed = seed,
                         n1=n1,
                         prop_null=prop_null,
                         pval = fl_pval$pval,
                         true_vec = fl_beta$beta != 0,
                         stringsAsFactors = F)
}
res_merge <- do.call(rbind, res)

saveRDS(res_merge, file = "output/gtex_type1.Rmd/res_merge.rds")
```



## Analysis


```{r}
res_merge <- readRDS(file = "output/gtex_type1.Rmd/res_merge.rds")

make_plots <- function(res, alpha, labels,
                       args=list(n1, labels)) {
  n_methods <- length(unique(res$method))
  cols <- RColorBrewer::brewer.pal(n_methods,name="Dark2")
  res %>% filter(n1==args$n1) %>% 
    group_by(method, seed) %>%
    summarise(type1=mean(pval<alpha, na.rm=T), nvalid=sum(!is.na(pval))) %>%
    ggplot(., aes(x=method, y=type1, col=method)) +
        # geom_errorbar(aes(ymin=mn+se, ymax=mn-se), width=.3) + 
        geom_boxplot() + geom_point(size=.7) + xlab("") +
      ylab("Type I error") +
      scale_x_discrete(position = "top",
                       labels=args$labels) +
      scale_color_manual(values=cols) +
      theme(axis.text.x=element_text(angle = 20, vjust = -.3, hjust=-.1))
}

library(cowplot)
levels(factor(res_merge$method))
labels <- c("deseq2", "edger", "limma_v", "sva_ttest", "t_test", "t_test_log2cpm_q", "wilcoxon")
  
make_plots(subset(res_merge, prop_null==1), alpha=.001,
             args=list(n1=50, labels=labels)) + 
            ggtitle("Type error at alpha < .001, 50/group") + 
            geom_hline(yintercept=.001, col="gray30", lty=3) +
            stat_summary(fun.y=median, geom="point", shape=18, size=6, col="black") +
            stat_summary(fun.y=mean, geom="point", shape=4, size=4, col="black")
```

> wilcoxon type I error is ~.06 for one dataset, and the corresponding type I error of t-test is ~.01, but for limma_voom is .001. Below I go over this null dataset. For 54 genes in this null dataset, wilcoxon test returned a smaller p-value than t-test. I investigated possible explanations for this, such as number of tied values and mean-variance relationship. But haven't reached a clear idea of why this may be the case? 

 
```{r check_outlier, eval=F}
# strange outlier
res <- subset(res_merge, prop_null==1);alpha=.001;args=list(n1=50)
out <- res %>% filter(n1==args$n1) %>% 
  group_by(method, seed) %>%
  summarise(type1=mean(pval<alpha, na.rm=T), nvalid=sum(!is.na(pval)))
out[which(out$type1 > .06),]
  
oo <- subset(res_merge, prop_null==1 & seed==93 & n1==50)
oo %>% group_by(method) %>%
  summarise(type1=mean(pval<alpha, na.rm=T), nvalid=sum(!is.na(pval)))
methods_vec <- unique(oo$method)
oo_print <- lapply(1:length(methods_vec), function(i) {
  which(oo[oo$method == methods_vec[i],]$pval < .001)
})
names(oo_print) <- methods_vec

# all sig. in wilcoxn also sig in t.test
setdiff(oo_print$t_test, oo_print$wilcoxon)

# genes sig. in wilcox but not in t.test
setdiff(oo_print$wilcoxon, oo_print$t_test)

# get the expression file
ff <- subset(dsc_res, method=="wilcoxon" & data_poisthin_gtex.prop_null==1 & data_poisthin_gtex.seed==93 & data_poisthin_gtex.n1==50)

df <- readRDS(file.path(dir_dsc, 
                        paste0(ff$data_poisthin_gtex.output.file, ".rds")))
check_genes <- setdiff(oo_print$wilcoxon, oo_print$t_test)

do.call(rbind, lapply(1:length(check_genes), function(i) {
  list(pval_wil=wilcox.test(df$Y[check_genes[i],]~df$X[,2], correct=T)$p.value,
       pval_t=t.test(log2(df$Y[check_genes[i],]+1)~df$X[,2])$p.value) } ) )

# check if the issue is related to ties in count data
# no...
dd <- sapply(1:nrow(df$Y), function(i) sum(duplicated(df$Y[i,])))
table(dd[oo_print$wilcoxon])

# check if the issue is related to mean-variance dependency
col_vec <- rep("black", nrow(df$Y))
col_vec[check_genes] <- "red"
v <- voom(df$Y, design=df$X, plot=T, save.plot = T)
plot(x=v$voom.xy$x,y=v$voom.xy$y,col=col_vec)
```


log2 scale by method by sample size

```{r}
make_plots_log2 <- function(res, alpha, labels,
                       args=list(n1, labels)) {
  n_methods <- length(unique(res$method))
  cols <- RColorBrewer::brewer.pal(n_methods,name="Dark2")
  res_plot <- res %>% filter(n1==args$n1) %>% 
    group_by(method, seed) %>%
    summarise(type1=mean(pval<alpha, na.rm=T), nvalid=sum(!is.na(pval))) 
  
  res_plot_mn <- res_plot %>% group_by(method) %>%
    summarise(mn=mean(type1, na.rm=T),
              med=median(type1, na.rm=T))
    # summarise(mn=mean(type1, na.rm=T), 
    #           n=sum(!is.na(type1)), se=sd(type1, na.rm=T)/sqrt(n)) %>%
    ggplot(data=res_plot, aes(x=method, y=log2(type1), col=method)) +
        # geom_errorbar(aes(ymin=mn+se, ymax=mn-se), width=.3) + 
        #geom_boxplot() + 
      geom_point(size=.7) + xlab("") +
      scale_x_discrete(position = "top",
                       labels=args$labels) +
      scale_color_manual(values=cols) +
      theme(axis.text.x=element_text(angle = 20, vjust = -.3, hjust=-.1)) +
        geom_point(data=res_plot_mn,
                   aes(x=method, y=log2(mn)), shape=4, size=4, col="black") +
        geom_point(data=res_plot_mn,
                   aes(x=method, y=log2(med)), shape=18, size=6, col="black")
}

library(cowplot)
levels(factor(res_merge$method))
labels <- c("deseq2", "edger", "limma_v", "sva_ttest", "t_test", "t_log2cpm_q", "wilcoxon")

make_plots_log2(subset(res_merge, prop_null==1), alpha=.001,
                 args=list(n1=5, labels=labels)) + 
                ggtitle("Type I error at alpha < .001, 5/group") + ylim(-11,-3) +
                geom_hline(yintercept=log2(.001), col="gray30", lty=3) +
                ylab("log2 type I error") 
make_plots_log2(subset(res_merge, prop_null==1), alpha=.001,
                 args=list(n1=10, labels=labels)) + 
                ggtitle("Type I error at alpha < .001, 10/group") + ylim(-11,-3) +
                geom_hline(yintercept=log2(.001), col="gray30", lty=3) +
                ylab("log2 type I error") 
make_plots_log2(subset(res_merge, prop_null==1), alpha=.001,
                 args=list(n1=50, labels=labels)) + 
                ggtitle("Type I error at alpha < .001, 50/group") + ylim(-11,-3) +
                geom_hline(yintercept=log2(.001), col="gray30", lty=3) +
                ylab("log2 type I error") 
make_plots_log2(subset(res_merge, prop_null==1), alpha=.001,
                 args=list(n1=150, labels=labels)) + 
                ggtitle("Type I error at alpha < .001, 150/group") + ylim(-11,-3) +
                geom_hline(yintercept=log2(.001), col="gray30", lty=3) +
                ylab("log2 type I error") 
```


log2 scale by sample size by method

```{r}
make_plots_log2_v2 <- function(res, alpha) {
  n_methods <- length(unique(res$method))
  cols <- RColorBrewer::brewer.pal(n_methods,name="Dark2")
  res_plot <- res %>% #filter(n1==args$n1) %>% 
    group_by(n1, method, seed) %>%
    summarise(type1=mean(pval<alpha, na.rm=T), nvalid=sum(!is.na(pval))) 
  res_plot$n1 <- factor(res_plot$n1)
  
  res_plot_mn <- res_plot %>% group_by(n1, method) %>%
    summarise(mn=mean(type1, na.rm=T),
              med=median(type1, na.rm=T))
  
  ggplot(data=res_plot, aes(x=n1, y=log2(type1), col=method)) +
    geom_point(size=.7) + 
    facet_wrap(~method) + 
    geom_point(data=res_plot_mn,
         aes(x=n1, y=log2(mn)), shape=4, size=3, col="black") +
    geom_point(data=res_plot_mn,
         aes(x=n1, y=log2(med)), shape=18, size=3, col="black") + #+ xlab("") +
    scale_color_manual(values=cols) +
    geom_hline(yintercept=log2(.001), col="gray30", lty=3)  +
    ylab("log2 Type I error") + xlab("sample size/group")
}

# library(cowplot)
# levels(factor(res_merge$method))
# labels <- c("deseq2", "edger", "limma_v", "t_test", "t_log2cpm_q", "wilcoxon")

make_plots_log2_v2(subset(res_merge, prop_null==1), alpha=.001) + 
                ggtitle("Type I error at alpha < .001") + ylim(-12,-3)
```


histogram of unadjusted p-value of one dataset

```{r, fig.width=10}
tmp <- subset(res_merge, prop_null==1 & n1==150) %>%
  group_by(seed, method) %>%
  summarise(type1=mean(pval < .001, na.rm=T)) 
tmp[which.max(tmp$type1),]
tmp[tmp$seed==89,]

subset(res_merge, prop_null==1 & n1==50 & seed==89) %>%
  ggplot(., aes(x=pval)) +
  geom_histogram(bins=30) +
  facet_wrap(~method)
```

