---
title: "Curate PBMC data used for type I error assessment in Van den Berg et al. 2018"
author: "Joyce Hsiao"
date: "2018-12-11"
output: workflowr::wflow_html
---

```{r options, echo=FALSE, results="hide",message=FALSE, error=FALSE, include=FALSE, autodep=TRUE, warning=FALSE}
knitr::opts_chunk$set(fig.align="center", error=FALSE, message=FALSE, warning=FALSE)
```

## Introduction


* Reproduce the experimental data used in Van den Berge et al., 2018 for creating PBMC null datasets. 

* Data includes 2,638 samples and 13,713 genes.


---

## Steps

Same steps as in `https://github.com/statOmics/zinbwaveZinger/blob/master/realdata/createdata/createDataObject.Rmd`. 


Reading in data.

```{r init, message=FALSE, eval=F}
library(Seurat)
library(dplyr)
library(Matrix)
# Load the PBMC dataset
pbmc.data <- Read10X(data.dir = "/project/mstephens/data/external_public_supp/pbmc3k_filtered_gene_bc_matrices")

pbmc <- CreateSeuratObject(raw.data = pbmc.data, min.cells = 3, min.genes = 200, project = "10X_PBMC")
```



QC and selecting cells for further analysis

```{r qc, results='hide', fig.height=7,fig.width=13, eval=F}
mito.genes <- grep(pattern = "^MT-", x = rownames(x = pbmc@data), value = TRUE)
percent.mito <- Matrix::colSums(pbmc@raw.data[mito.genes, ]) / Matrix::colSums(pbmc@raw.data)
# AddMetaData adds columns to object@meta.data, and is a great place to stash QC stats
pbmc <- AddMetaData(object = pbmc, metadata = percent.mito, col.name = "percent.mito")

pbmc <- FilterCells(object = pbmc, subset.names = c("nGene", "percent.mito"), low.thresholds = c(200, -Inf), high.thresholds = c(2500, 0.05))
```

Normalizing the data

```{r normalize, results='hide', eval=F}
pbmc <- NormalizeData(object = pbmc, normalization.method = "LogNormalize", scale.factor = 1e4)
```

Detection of variable genes across the single cells

```{r var_genes, fig.height=7, fig.width=11, results='hide', eval=F}
pbmc <- FindVariableGenes(object = pbmc, mean.function = ExpMean, dispersion.function = LogVMR, x.low.cutoff = 0.0125, x.high.cutoff = 3, y.cutoff = 0.5)

length(x = pbmc@var.genes)
```

Scaling the data and removing unwanted sources of variation


```{r regress, fig.height=7, fig.width=11, results='hide', eval=F}
pbmc <- ScaleData(object = pbmc, vars.to.regress = c("nUMI", "percent.mito"))
```

Perform linear dimensional reduction

```{r pca, eval=F}
pbmc <- RunPCA(object = pbmc, pc.genes = pbmc@var.genes, do.print = TRUE, pcs.print = 1:5, genes.print = 5)
```

Cluster the cells

```{r cluster, fig.height=5, fig.width=7, eval=F}
pbmc <- FindClusters(object = pbmc, reduction.type = "pca", dims.use = 1:10, resolution = 0.6, print.output = 0, save.SNN = TRUE)
```


Run Non-linear dimensional reduction (tSNE)

```{r tsne, fig.height=5, fig.width=7, message=FALSE, warnings=FALSE, eval=F}
pbmc <- RunTSNE(object = pbmc, dims.use = 1:10, do.fast = TRUE)
```


Assigning cell type identity to clusters

Cluster ID | Markers       | Cell Type
-----------|---------------|----------
0          | IL7R          | CD4 T cells
1          | CD14, LYZ     | CD14+ Monocytes
2          | MS4A1         | B cells
3          | CD8A          | CD8 T cells
4          | FCGR3A, MS4A7 | FCGR3A+ Monocytes
5          | GNLY, NKG7    | NK cells
6          | FCER1A, CST3  | Dendritic Cells
7          | PPBP          | Megakaryocytes


```{r labelplot, fig.height=5, fig.width=9, warning = FALSE, eval=F}
current.cluster.ids <- c(0, 1, 2, 3, 4, 5, 6, 7)
new.cluster.ids <- c("CD4 T cells", "CD14+ Monocytes", "B cells", "CD8 T cells", "FCGR3A+ Monocytes", "NK cells", "Dendritic cells", "Megakaryocytes")
pbmc@ident <- plyr::mapvalues(x = pbmc@ident, from = current.cluster.ids, to = new.cluster.ids)
```

Further subdivisions within cell types

```{r investigate_split, fig.width=15, eval=F}
# First lets stash our identities for later
pbmc <- StashIdent(object = pbmc, save.name = "ClusterNames_0.6")
# Note that if you set save.snn=T above, you don't need to recalculate the SNN, and can simply put: 
# pbmc <- FindClusters(pbmc,resolution = 0.8)
pbmc <- FindClusters(object = pbmc, reduction.type = "pca", dims.use = 1:10, resolution = 0.8, print.output = FALSE)
```


Create SE object

```{r, eval=F}
library(SingleCellExperiment)
# use raw data as input for zinbwave but keep only non filtered cells
# and most variable genes indentified by seurat
keepcells = colnames(pbmc@data)
counts = pbmc@raw.data[, keepcells]
# zinbwave does not want dgTMatrix as input
counts = as.matrix(counts)
counts = counts[rowSums(counts) > 0, ]
#keepcells = as.integer(pbmc@ident) %in% 1:2
#counts = counts[, keepcells]
# coldata
clusters = as.integer(pbmc@ident)
#clusters = clusters[clusters %in% 1:2]
cData = data.frame(seurat = clusters)
rownames(cData) = colnames(counts)
# rowdata
rData = data.frame(seuratVarGenes = rownames(counts) %in% pbmc@var.genes)
rownames(rData) = rownames(counts)
# create sce object
core = SingleCellExperiment(assays = list(counts = counts),
                            colData = cData, rowData = rData)

unloadNamespace("Seurat")
saveRDS(core, file = '../dsc/data/pbmc.rds')

saveRDS(assay(pbmc), file = '../dsc/data/pbmc_counts.rds')
```


## Outputting objects

All filtered samples

```{r, eval=F}
pbmc <- readRDS(file = '../dsc/data/pbmc.rds')

pbmc
```


Samples from cluster 1 and 2

```{r, eval=F}
pbmc <- readRDS(file = '../dsc/data/pbmc.rds')

pbmc_counts_sub <- assay(pbmc[,colData(pbmc)$seurat==1 | colData(pbmc)$seurat==2 ] )
saveRDS(pbmc_counts_sub, file = '../dsc/data/pbmc_counts_sub.rds')
```

