---
title: "Evaluate two-sample tests"
author: "jhsiao999"
date: "2018-11-15"
output: workflowr::wflow_html
---

## Overview

We applied two-sample tests for mean difference to simulated datasets.

The methods applied include:

* DESeq2
* glm poisson
* glm quasipoisson
* t-test
* wilcoson tet

The simulated data were made from experimental datasets as described below. 

1. Draw (sample without replacement) p genes and N samples from the experimental dataset. 

2. For the `Correlated case`, assign samples to Group 1 or Group 2. 

3. For the `Uncorrelated case`, permute sample labels for each gene, then in the permuted sample, assign samples to Group 1 or Group 2. 

In addition, the parameter setting is:

* Sample size (n1 vs n2): 50 vs 50, 300 vs 300

* Number of genes (p): 946 

Regarding the experimental data:

* single cell RNA-seq data from 10x genomics technology

* The dataset used in the simulation include CD14+ Monocytes and CD8 T cells, a total of 787 samples and 946 genes. 

* The original complete dataset include 2,700 single cells and 32,738 genes from 8 immune cell types: B cells, CD4 T cells, CD8 T cells, CD14+ Monocytes, Dendritic cells, FCGR3A+ Monocytes, Megakaryocytes, NK cells. 

* We filtered the original data to include samples that are detected in > 200 genes and genes that are detected in > 20% of cells.


---

## Results

```{r, warning=F, message=F}
library(dscrutils)
#setwd("~/Dropbox/GitHub/dsc-log-fold-change/dsc")
dir_dsc <- "/Users/joycehsiao/Dropbox/GitHub/dsc-log-fold-change/dsc/benchmark"
out <- dscquery(dir_dsc, 
                c("get_data", "get_data.n1", "get_data.n2", "method", "method.p"))


plotRest <- function(dir_dsc, dscoutput, 
                     sim_case=c("sample_correlated", "sample_uncorrelated"), 
                     sample_size=c(50,300),
                     title_label, plot=T, return_res=T) {
  n_methods <- nlevels(dscoutput$method)
#dscoutput <- out; sim_case="random_sample"; sample_size=50  
  out.sub <- dscoutput[dscoutput$get_data==sim_case & dscoutput$get_data.n1==sample_size,]
  
  res <- vector("list",n_methods)
  for (i in 1:nrow(out.sub)) {
#    print(i)
    fl <- readRDS(file.path(dir_dsc,
                         paste0(as.character(out.sub$method.p[i]), ".rds")))
    res[[i]] <- data.frame(method = as.character(out.sub$method)[i],
               n1_n2 = paste0(out.sub$get_data.n1[i],".",
                              out.sub$get_data.n2[i]),
               pval = fl$p,
               stringsAsFactors = F)
  }
  names(res) <- as.character(out.sub$method)
  
  if (plot) {
    cols <- RColorBrewer::brewer.pal(n_methods,name="Accent")
    par(mfrow=c(3,2))
    for (i in 1:length(res)) {
      hist(res[[i]]$pval, main="",
           xlab = "p-values", ylab = "Frequency",
           nclass = 20, col=cols[i])
      title(main=names(res)[i])
    }
    title(title_label, outer=T, line=-1)
    
    qq <- lapply(1:length(res), function(i) {
      qqplot(x=runif(sample_size*2,0,1), y=res[[i]]$pval, plot.it=F)
    })
    plot(qq[[1]]$x, qq[[1]]$y, col = cols[1], cex=.7, pch = 16,
         xlab = "Uniform(0,1)", ylab = "Empirical distribution",
         main = "QQ-plot")
    for (i in 2:n_methods) {
      points(qq[[i]]$x, qq[[i]]$y, col = cols[i], cex=.7, pch = 16)
    }
    abline(0,1, col = "black")
    title(title_label, outer=TRUE, line=-1)
  }
  
  if (return_res) {
  return(res)
  }
}
```

```{r, fig.height=7}
out.corr.50 <- plotRest(dir_dsc=dir_dsc, dscoutput=out, 
                         sim_case="sample_correlated", 
                         sample_size=50,
                         title_label="Correlated case: 50 vs 50", plot=T, return_res=T)

out.corr.300 <- plotRest(dir_dsc=dir_dsc, dscoutput=out, 
                         sim_case="sample_correlated", 
                         sample_size=300,
                         title_label="Correlated case: 300 vs 300", plot=T, return_res=T)

out.uncorr.50 <- plotRest(dir_dsc=dir_dsc, dscoutput=out, 
                         sim_case="sample_uncorrelated", 
                         sample_size=50,
                         title_label="Uncorrelated case: 50 vs 50", plot=T, return_res=T)

out.uncorr.300 <- plotRest(dir_dsc=dir_dsc, dscoutput=out, 
                         sim_case="sample_uncorrelated", 
                         sample_size=300,
                         title_label="Uncorrelated case: 300 vs 300", plot=T, return_res=T)
```

